---
title: "Mapping overdoses in Cincinnati"
output: html_notebook
---

Today we will analyze data from the Cincinnati Fire Incidents database. The database is publicly available and can be downloaded here: https://data.cincinnati-oh.gov/Safer-Streets/Cincinnati-Fire-Incidents-CAD-including-EMS-ALS-BL/vnsz-a3wp

The city also has a nice graphical interface for mapping and exploring these data. https://insights.cincinnati-oh.gov/stories/s/Heroin-Overdose-Responses/dm3s-ep3u/

These data contain a record of emergency medical service calls, including fires, rescues and other fire department-related services. Our goal will be to examine spatial and spatiotemporal clusters

```{r setup, include=F}
#install.packages('rgdal')
#install.packages('raster')

library(lubridate)
library(rgdal)
library(raster)
library(surveillance)
```


```{r}

cin1 <- read.csv('https://data.cincinnati-oh.gov/api/views/vnsz-a3wp/rows.csv?accessType=DOWNLOAD')
cin1<-readRDS('./Data/cincinnati.rds')
```

## Let's first have a look at what variables are in the dataset
```{r}
names(cin1)
```

```{r}
View(cin1)
```


What are the different types of calls in the database?
```{r}
sort(
  table(
    cin1$CFD_INCIDENT_TYPE_GROUP #what variable are we tabulating?
    ), 
  decreasing=T) #biggest to smallest sort
```


## Format the date and time variables
```{r}
cin1$CREATE_TIME_INCIDENT<-as.character(cin1$CREATE_TIME_INCIDENT) #convert date-time to a character vector
```

```{r}
date_time_list<-    strsplit(cin1$CREATE_TIME_INCIDENT,' ')#split date and time into separate vectors, this is saved as a list

dtparts = do.call(rbind,date_time_list) #combine pieces of list back into a data frame    
head(dtparts)
```

```{r}
cin1$date<-as.Date(dtparts[,1], '%m/%d/%Y') #tell R this is a date with format month/day/year; add it to the cin1 data frame
```

```{r}
cin1$weekdate<-floor_date(cin1$date, unit='week') #round date down to nearest Sunday, add it to the cin1 data frame
```

Pull out the hour of the incidence
```{r}
cin1$time<-dtparts[,2]
cin1$hour<-as.numeric(gsub(":.*", "",cin1$time)) #Extract out the hour of the call
```

## Let's subset to just the overdose data 
```{r}
op1<-cin1[cin1$CFD_INCIDENT_TYPE_GROUP=='HEROIN OVERDOSE',] #pull out rows where incident type is heroin overdose
```

#Also subset out breathing problem complaints
We will use this later on for our SATSCAN analysis...
```{r}
br1<-cin1[cin1$CFD_INCIDENT_TYPE_GROUP=='BREATHING PROBLEMS',]

```

## distribution of the dates
```{r}
hist(op1$date, breaks=98) #histogram with 98 bins
```

#Look at distribution of overdoses over time
```{r}
hist(op1$hour, breaks=24, freq=T) #histogram with 24 breaks. x axis shows hours from 12:01am-11:59pm
```

## Simple maps
What we will do here is just make an X-Y plot with latitude on the y axis and longitude on the x axis. For each overdose, we will make a dot. 

```{r, fig.height=5, fig.width=5}
plot(op1$LONGITUDE_X, #x variable
     op1$LATITUDE_X , #Y variable
     pch=16,  #shape of the markers
     cex=0.5, #size of the marker
     col='black',  #specify a black marker
     bty='l')
```

This is hard to see because there are so many dots. Let's make them partially transparent so that we can see where they are piling up. 
```{r, fig.height=5, fig.width=5}
col.plot=rgb(1,0,0,alpha=0.02) #red, with transparency
plot(op1$LONGITUDE_X, #x variable
     op1$LATITUDE_X , #Y variable
     pch=16,  #shape of the markers
     cex=0.5, #size of the marker
     col=col.plot,  #specify a transparent red
     bty='l')
```
#Lets add in neighborhood boundaries
This shows a 'hotspot' in downtown Cincinnatti. Even without doing any fancy statistics we can start to see some patterns

Read in a file that has map coordinates for neighborhood boundaries
```{r}
shp.cin<-readRDS( 'cincinnati_neighborhood.map.rds')

```


```{r}
plot(shp.cin, bty='l')
col.plot=rgb(1,0,0,alpha=0.02) #red, with transparency
points(op1$LONGITUDE_X, #x variable
     op1$LATITUDE_X , #Y variable
     pch=16,  #shape of the markers
     cex=0.5, #size of the marker
     col=col.plot  #specify a transparent red
    )
```

## As a comparison, let's look at breathing issues
From this we can see some similarities and some differences with the opiod map
```{r}
plot(shp.cin, bty='l')
col.plot=rgb(0,0,1,alpha=0.02) #blue, with transparency
points(br1$LONGITUDE_X, #x variable
     br1$LATITUDE_X , #Y variable
     pch=16,  #shape of the markers
     cex=0.5, #size of the marker
     col=col.plot  #specify a transparent red
    )
```


## Prepare data for SATSCAN to detect spatiotemporal clusters
We will run 3 analyses in SATSCAN: 
1. A purely temporal analysis to look for temporal clusters 
2. A space-time permutation analysis to look for space-time cluster
3.A purely spatial analysis using breathing conditions as controls and heroin overdoses as cases. For the purely spatial analysis, because we are using point data, we don't have a population denominator, so a trick is to use a different event (breathing problems) to adjust for spatial variation in population. We perform a 'casecontrol' analysis where we are evaluating the probability of detecting a case vs control in each of the circles.
What are the advantages and disadvantages of this?

1)Create an identifier for each event This will just be an index variable. We will combine together the heroin overdoses and breathing complaints into a single data frame so that we have unique IDs for all cases and controls.
```{r}
all<-rbind.data.frame(op1, br1)
all$id<-1:nrow(all)
```

1a) Remove rows withut coordinate data
```{r}
all.nomiss<-all[!is.na(all$LATITUDE_X),]
```


2a) create a case file
```{r}
op.case<-all.nomiss[all.nomiss$CFD_INCIDENT_TYPE_GROUP=='HEROIN OVERDOSE',c('id','date')]
write.csv(op.case,'op.case.csv')
```

2b) create a control file
```{r}
br.case<-all.nomiss[all.nomiss$CFD_INCIDENT_TYPE_GROUP=='BREATHING PROBLEMS',c('id','date')]
write.csv(br.case,'br.control.csv')
```

3) create a geo file
```{r}
all.geo<-all.nomiss[,c('id',"LATITUDE_X", "LONGITUDE_X"  )]
write.csv(all.geo,'all.geo.csv')

#output just the opiod coordinates
op.geo<-all.nomiss[all.nomiss$CFD_INCIDENT_TYPE_GROUP=='HEROIN OVERDOSE',c('id',"LATITUDE_X", "LONGITUDE_X"  )]
write.csv(op.geo,'op.geo.csv')
```
4) Go to SATSCAN and run a space-time permutation analysis
--Run a purely temporal analysis with op.geo.csv and op.case.csv
--Run a space-time analysis with op.geo.csv and op.case.csv
---start date 7-17-2016; end date April 8, 2019 
--For computation efficiency, might want to change number of simulations from 999 to 99 (analysis tab/advanced options/inference)
--(Run a Bernoulli case-control analysis with the breathing conditions as controls--pre-run for you because it takes forever... use all.geo.csv, br.control.csv, op.geo.csv)

5a) See results of your space-time permutation analysis here:
Remember that these are space-time clusters, so hey occur for a distinct point in time, so might not line up perfectly with areas with high incidence on average.
```{r}
space.time.shp<- shapefile('./cincinnatti od/cin space time.col.shp')
plot(space.time.shp, border='blue') #Plot the circles around cluster
plot(shp.cin,add=T) #Add the borders of neighborhoods
col.plot=rgb(1,0,0,alpha=0.02) #blue, with transparency
#Then add the case data
points(op1$LONGITUDE_X, #x variable
     op1$LATITUDE_X , #Y variable
     pch=16,  #shape of the markers
     cex=0.5, #size of the marker
     col=col.plot  #specify a transparent red
    )
```

5b) See results of the pre-run purely spatial case-control analysis here:
```{r}
case.control.shp<- shapefile('./cincinnatti od/cin1.col.shp')
plot(case.control.shp, border='blue') #Plot the circles around cluster
plot(shp.cin,add=T) #Add the borders of neighborhoods
col.plot=rgb(1,0,0,alpha=0.02) #blue, with transparency
#Then add the case data
points(op1$LONGITUDE_X, #x variable
     op1$LATITUDE_X , #Y variable
     pch=16,  #shape of the markers
     cex=0.5, #size of the marker
     col=col.plot  #specify a transparent red
    )
```

## Let's also set up a CUSUM algorithm for rapid detection of overdoses

```{r}
op1$count=1
op2<- aggregate(op1$count, by=list('date'=op1$date), FUN=sum) #aggregate by day
#hist(cusum1$x)
```

Fill time series to ensure all days represented
```{r}
dates.present<-unique(op2$date)
all.days<- cbind.data.frame('date'=seq.Date(from=dates.present[1], to=dates.present[length(dates.present)], by='day'))
```

merge that full list of days with original data
```{r}
op2<-merge(op2, all.days, by='date')
op2$x[is.na(op2$date)]<-NA
```


How many overdoses per day?
```{r}
hist(op2$x)
median(op2$x)
```
create a surveillance object
```{r}
op_DP <- create.disProg(
      week = 1:nrow(op2), #Index of observations
      observed = op2$x ,
      state=matrix(0, nrow=nrow(op2), ncol=1),
      start = c(2015, 26))
```

```{r}
time.points.test<-48:nrow(op2)
```

```{r}
in.control.mean<-5
out.of.control.mean<- in.control.mean*2

k.set= (in.control.mean-out.of.control.mean )/(log(in.control.mean)-log(out.of.control.mean))
k.set.std<- (k.set- in.control.mean)/sqrt(in.control.mean)
h.set=4

```

```{r}
cusum1<- algo.cusum(op_DP,
    control=list(
      range=time.points.test,
   k=k.set,
   h=h.set) 
)
plot(cusum1)
```

##Plot the CUSUM output
```{r, fig.width=5, fig.height=10}
#PLOT THE OUTPUT
par(mfrow=c(3,1))
plot(op2$x, main='Full time series', bty='l')
abline(h= k.set )

std.outcome<- (op2$x-mean(op2$x))/sqrt(mean(op2$x))
plot(std.outcome[time.points.test], bty='l',type='l')
abline(h= k.set.std )
title(paste0('Observed data with K=',round(k.set,2)))

plot(cusum1$cusum, type='p', bty='l',pch=16, col=cusum1$alarm+1)
abline(h= h.set)
title(paste0('CUSUM statistic with H=', round(h.set,2),' red=ALARM'))

```

